Computing properties of independent sets by generic programming tensor networks \thanks{\funding{...}}

### 2022-03-18 paper/paper.tex
Updates from Overleaf

\begin{abstract}
%We introduce a method that uses generic programming tensor networks to compute various properties of a class of hard graph problems that can be decribed by local constraints.
We introduce a tensor network algorithm %a method that combines generic programming and tensor network techniques 
to compute various solution space properties of combinatorial optimization problems on graphs with local constraints. \purple{I still think there is a more concise way to word the class of combinatorial optimization problems this works on...}
%Let the independent set problem be an example, we show how to compute, for example, the maximum several independent set sizes, the counting of independent sets of a give size, the enumeration of independent sets of a given size.
We look at the independent set problem as an example, and show how to compute the size of the maximum independent set, count the number of independent sets of a given size, and enumerate the independent sets of a given size.
%By making use of generic programming, our algorithms are very simple to implement and one can in addition directly utilize recent advances in tensor network contraction techniques such as near-optimal contraction order finding and slicing to achieve high performance.
By using generic programming techniques, %which require a program to work correctly for different input data types without sacrificing efficiency,
the same general, simple-to-implement framework can be used to compute all of these properties. Our algorithm utilizes recent advances in tensor network contraction techniques to achieve high performance, including methods to quickly find a near-optimal contraction order and slicing.
The algorithmic complexity of this approach is $2^{{\rm tw}(G)}$, where ${\rm tw} (G)$ is the treewidth of the problem graph $G$.
Other hard graph problems solvable in our framework include the maximum cut problem, the vertex coloring problem, the maximal clique problem, the dominating set problem, and the satisfiability problem, among others.
To demonstrate the versatility of this tool, we apply it to a few examples, including computing the entropy constant for several hardcore gases on disordered lattices, studying the overlap gap property on 3-regular graphs, and visualizing the output of quantum optimization algorithms for the independent set problem. % including the calculations of the entropy constant for some hardcore lattice gases on 2D square lattices and the study of the overlap gap property on three regular graphs.
We show how our method helps to understand these hard problems better. \purple{Say on systems with hundreds of vertices or spins?}
\end{abstract}

@@

The most studied aspect of the above problems is finding an optimal solution, 
%The most studied version of the above problems is finding an optimal solution maximum/minimum set satisfying the contraints and the corresponding set size,
, which is typically 
%while many of them belong to the hardness category
NP-hard~\cite{Hastad1996}, i.e. it is unlikely to be solved in a polynomial time.
%\purple{Is MIS really the most studied NP-hard optimization problem? I would guess it is heavily studied, but not necessarily the most studied.}
%\blue{Here, we haven't mentioned the independent set, here ``version'' means, e.g. counting, enumeration, devision. I think finding one of the largest/smallest set and its cardinality in the above mentioned problems is the most studied?}
In this paper, we focus on the less studied problem of computing \textit{solution space properties}  which can be much harder, e.g. counting independent sets of a general graph is \#P-complete~\cite{Vadhan2001}, but is crucial towards understanding a hard problem.
Here, the \textit{solution space property} refers to a class of quantities that not only include the maximum/minimum set size,
but also include the number of sets at a give size, enumeration of all sets at a give size or direct sampling of such sets when they are not enumeratable.
For the weighted version of the above problems, it also includes finding largest/smallest $k$ sets and their sizes.
%are many interesting and hard computational problems concerning various properties of independent sets.
Solution space property computation allows us to understand the solution space better.
With the counting of sets at different sizes, one can related the computational hardness with the counting of sets at different sizes~\cite{Xu2018}.
With the configuration enumeration/sampling at a given size, one can measure the overlap gap property better,
while the configuration space connectivity also privides us an intuition about designing a better classical or quantum algortihms.
\purple{I would give examples of why we want to compute solution space
properties.
For instance, in statistical physics, computing thermal states and
observables via the partition function. In complexity theory, proving limits on local
algorithms (OGP).}\blue{now?}

@@

To avoid the discussion being unnecessarily diversed, we mainly focus on the independent set problem,
and discuss other combinatorial optimization problems,
including the cutting problem, the matching problem, the vertex coloring problem, the satisfiability problem, the dominating set problem, the set packing problem and the maximal clique problem, in \App{sec:otherproblems}.
%\xpurple{More details on which combinatorial optimization problems this applies to?}
The paper is organized as follows.
We first introduce the basic concepts of tensor networks and generic programming in \Sec{sec:tn} and \Sec{sec:generic}.
Then we show how to reduce the independent set problem to a tensor network contraction problem in \Sec{sec:tnmap}
and show how to engineer the element types in \Sec{sec:counting}, \Sec{sec:enumeration} and \Sec{sec:weighted} to count independent sets at a given size, enumerate/sample independent sets at a given size and find largest set sizes in weighted graphs.
% by computing a number of independent set properties of certain sparse graphs on central processing units (CPUs) and graphics processing units (GPUs) and show its good performance.
Lastly, we benchmark our algorithms in \Sec{sec:benchmark} and provide a few examples in \Sec{sec:examples} to demonstrate the versatility of our tool.
We show how our method can help computing the entropy constant for some hardcore lattice gases on 2D square lattices
and analyzing the presence of overlap gap property on three regular graphs.
%Our method can also be used to find ``maximal'' independent set properties; 
%a maximal independent set is an independent set that is not a subset of any other independent set, but its size may not be the maximum.

@@

\begin{definition}[Generic programming ~\cite{Stepanov2014}]
   Generic programming is an approach to programming that focuses on designing algorithms and data structures so that they work in the most general setting without loss of efficiency.
\end{definition}
\purple{efficiency}
By in the most general setting, we mean that a single program should work correctly for different input data types. For example, 
%\purple{I think the following paragraph could be clarified. When you read the definition of generic programming, it's not clear what ``in the most general setting'' means. My understanding is it just means that the program will work regardless of the input data type - it would be good to say this explicitly. }
%\blue{This definition is from the book I cited, I also do not like ``in the most general setting'', maybe we just use our version?}
%\purple{I also think the way it is written takes a while to understand what dynamically versus statically typed languages are. The definition, “When these languages “see” a new input type, the compiler can recompile the generic program for the new type.” comes rather late in the paragraph.}
%\xpurple{Also, while I know what you mean, concepts like “inlining immutable elements” and “cache miss rate” would be
%lost on someone only with a physics background.}
%This definition of generic programming contains two major aspects: a single program works in the most general setting and efficiency.
%To understand the first aspect on generality,
suppose we want to write a function that raises an element to a power, $f(x, n) := x^n$.
One can easily write a function for standard number types that computes the power of $x$ in $O \left( \log(n) \right)$ steps using the multiply and square trick.
Generic programming does not require $x$ to be a standard number type,
instead it treats $x$ as an element with an associative multiplication operation $\odot$ and a multiplicative identity $\mymathbb{1}$.
In such a way, when the program takes a matrix as an input, it computes the matrix power without extra efforts.
The second aspect is about performance. For dynamically typed languages, such as Python,
one can easily write very general codes, but the efficiency is not guaranteed; for example, the speed of computing the matrix multiplication between two numpy arrays with python objects as elements is much slower than statically typed languages such as C++ and Julia~\cite{Bezanson2012}.
C++ uses templates for generic programming while Julia takes advantage of just-in-time compilation and multiple dispatch.
When these languages ``see'' a new input type, the compiler can recompile the generic program for the new type.
A myriad of optimizations can be done during the compilation, such as optimize the memory layout of immutable elements with fixed sizes in an array to speed up the array indexing.
In Julia, arrays with immutable elements with fixed sizes can even be compiled to GPU devices for faster computation~\cite{Besard2018}.

This motivates us to think about what is the most general tensor element type allowed in a tensor network contraction program.
We find that, as long as the algebra of tensor elements forms a commutative semiring, the tensor network contraction result will be valid and be independent of the contraction order.
A commutative semiring is a ring with its multiplication operation being commutative and without an additive inverse.
To define a commutative semiring with the addition operation $\oplus$ and the multiplication operation $\odot$ on a set $R$, the following relations must hold for any arbitrary three elements $a, b, c \in R$.
\begin{align*}
(a \oplus b) \oplus c = a \oplus (b \oplus c) & \hspace{5em}\text{$\triangleright$ commutative monoid $\oplus$ with identity $\mymathbb{0}$}\\
a \oplus \mymathbb{0} = \mymathbb{0} \oplus a = a &\\
a \oplus b = b \oplus a &\\
&\\
(a \odot b) \odot c = a \odot (b \odot c)  &   \hspace{5em}\text{$\triangleright$ commutative monoid $\odot$ with identity $\mymathbb{1}$}\\
a \odot  \mymathbb{1} =  \mymathbb{1} \odot a = a &\\
a \odot b = b \odot a &\\
&\\
a \odot (b\oplus c) = a\odot b \oplus a\odot c  &  \hspace{5em}\text{$\triangleright$ left and right distributive}\\
(a\oplus b) \odot c = a\odot c \oplus b\odot c &\\
&\\
a \odot \mymathbb{0} = \mymathbb{0} \odot a = \mymathbb{0}
\end{align*}
We require the algebra being multiplicative commutative because the contraction order optimizer will change the multiplication order of elements, while we require that the contraction order does not change the contraction results.
In the following sections, we show how to compute a number of properties of independent sets by designing tensor element types as specific commutative semirings without changing the tensor network contraction program~\cite{Stepanov2014}.
The Table~\ref{tbl:generictypes} summarizes those properties that can be computed by various tensor element types.

\begin{table}[t!]\centering
\begin{minipage}{\columnwidth}
\ra{1.3}
        \begin{tabularx}{\textwidth}{sb}\toprule
            \hline
   \textbf{Element type}     & \textbf{Solution space property} \\
   {$\mathbbm{R}$}     & {Counting of all independent sets} \\
   {Polynomial} (\Eq{eq:polynomial})     & {Independence polynomial} \\
   {Tropical (\Eq{eq:tropical})}    & {Maximum independent set size} \\
   {Extended tropical of order $k$ (\Eq{eq:ext-tropical})}    & {Largest $k$ independent set sizes} \\
   {Polynomial truncated to $k$-th order (\Eq{eq:countingtropical} and \Eq{eq:max2poly})}     & {$k$ largest independent sizes and their degeneracy} \\
   {Set} (\Eq{eq:set})     & {Enumeration of independent sets} \\
   {Expression tree} (\Eq{eq:expr})     & {Sampling of independent sets} \\
   {Polynomial truncated to highest order combined with bit string} (\Eq{eq:singleconfig})     & {Maximum independent set size and one of such configurations} \\
   {Polynomial truncated to $k$-th order combined with set} (\Eq{eq:countingtropicalset})    & {$k$ largest independent set sizes and their enumeration} \\
            \bottomrule
        \end{tabularx}
    \caption{Tensor element types and the independent set properties that can be computed using them.}\label{tbl:generictypes}
\end{minipage}
\end{table}

@@

    The contraction of this tensor network can be done in a pairwise order utilizing the associativity, additive commutativity, and multiplicative commutativity of tensor elements:
    \small{
    \begin{align*}
        &\sum_{s_a,s_b,s_c,s_d,s_e} W(x_a, w_a)_{s_a} W(x_b, w_b)_{s_b} W(x_c, w_c)_{s_c} W(x_d, w_d)_{s_d} W(x_e, w_e)_{s_e} B_{s_a s_b} B_{s_b s_d} B_{s_c s_d} B_{s_a s_c} B_{s_b s_c} B_{s_d s_e}\\
        =&\sum_{s_b,s_c}\left(\sum_{s_d}\left(\left(\left(\left(\sum_{s_e}B_{s_d s_e}W(x_e,w_e)_{s_e}\right) W(x_d,w_d)_{s_d}\right) \left(B_{s_bs_d} W(x_b, w_b)_{s_b}\right)\right) \left(B_{s_cs_d} W(x_c, w_c)_{s_c}\right)\right)\right.\\
        &\phantom{XXX}\left.\left(B_{s_bs_c}\left(\sum_{s_a}B_{s_as_b}\left(B_{s_as_c}W(x_a,w_a)_{s_a}\right)\right)\right)\right)\\
        =&1 + x_a^{w_a} + x_b^{w_b} + x_c^{w_c} + x_d^{w_d} + x_e^{w_e} + x_a^{w_a}x_d^{w_d} + x_a^{w_a}x_e^{w_e} + x_c^{w_c}x_e^{w_e} + x_b^{w_b}x_e^{w_e}
        %=&1+5x+4x^2 \qquad \quad (x_{i} = x, w_{i}=1).
    \end{align*}}

One can easily check the maximum tensor during computing is rank 2, and the total number of multiplication operations is reduced significantly.
\end{example}

@@

and then apply the gaussian elimination to fit the coefficients of this polynomial.

@@

This algebra is a commutative semiring because we define the equivalence of two expression trees by comparing their expanded (using \Eq{eq:set}) forms, rather than their storages.
Although one is probably unable to enumerate the configurations represented by this tree due to the extremely large configuration space size,
but he can directly sample configurations from this tree.
This is because once we have this expression tree representation of the configuration space,
either generate by using this algebra directly or serve as the coefficients of (truncated) polynomials,
one can compute the number of configurations in each branch rigorously and efficiently.
%\purple{To clarify - is the expression tree now the coefficients
%on the polynomial? I am a little confused on how we would, for instance, use this to
%compute the MISs}\blue{added one more descriptive sentence, does it resolve your concern?}
To generate samples, one just descends this expression tree to left or right sibling with probability decided by the size of each branch while doing the computation specified by the first field.
\xpurple{How do we compute the size of each
branch?}
\begin{example}
Let us consider the following expression tree 
\begin{equation*}
    (\texttt{SUM}, /, (\texttt{PROD}, /, A, B), (\texttt{SUM}, /, C, D)),
\end{equation*}
where siblings $A, B, C$ and $D$ can be any four types of nodes.
It can be represented diagramatically as the following.

@@

We compare two categories of graphs, one is the King's graph at $0.8$ filling, a type of unit disk graph having no overlap gap property, \blue{@Maddie, do you have a good reference to justify this statement about unit disk graph OGP?}

@@

@@

\begin{figure}[th]
   \centering
\centerline{\begin{tikzpicture}[]
    \scriptsize
    \node[draw,fill=lime!80,fill opacity=1, text opacity=1.0,ellipse,minimum width=2cm, minimum height=1cm,inner sep=0pt] at (0, 2.5) (R) {$\mathbbm{R}$};
    \def\dx{-3};
    \node[draw,fill=teal!50,fill opacity=1, text opacity=1.0,ellipse,minimum width=5cm, minimum height=3cm,inner sep=0pt] at (\dx, 0) (PN) {\hspace{-3.5cm}Polynomial};
    \node[draw,fill=brown!75,fill opacity=1, text opacity=1.0,ellipse,minimum width=3.5cm, minimum height=1.0cm,inner sep=0pt] at (\dx, 0.7) (P1) {\hspace{-1.0cm}Largest order};
    \node[draw,fill=brown!40,fill opacity=1, text opacity=1.0,ellipse,minimum width=3.5cm, minimum height=1.0cm,inner sep=0pt] at (\dx, -0.4) (P2) {\hspace{-0.8cm}2nd largest order};
    \node[draw,fill=brown,fill opacity=1, text opacity=1.0,ellipse,minimum width=0.8cm, minimum height=0.3cm,inner sep=0pt] at (\dx, 1.0) (T) {};
    \node[draw,fill=brown!50,fill opacity=1, text opacity=1.0,ellipse,minimum width=0.8cm, minimum height=0.3cm,inner sep=0pt] at (\dx, -0.1) (T2) {};
    \node at (\dx, -1.2) {$\ldots$};
    \node[above = 1cm] at (T) (textT) {Tropical};
    \node[below = 2cm, left=4cm] (textT2) {Extended Tropical};
    \draw[black,-latex] (textT) -- (T);
    \draw[black,-latex] (textT2) -- (T2);

>>>>

\begin{abstract}
%We introduce a method that uses generic programming tensor networks to compute various properties of a class of hard graph problems that can be decribed by local constraints.
We introduce a tensor network algorithm %a method that combines generic programming and tensor network techniques 
to compute various solution space properties of a class of combinatorial optimization problems specified by local constraints. \purple{I still think there is a more concise way to word the class of combinatorial optimization problems this works on...}\blue{changed a bit, better?}
%Let the independent set problem be an example, we show how to compute, for example, the maximum several independent set sizes, the counting of independent sets of a give size, the enumeration of independent sets of a given size.
We look at the independent set problem as an example, and show how to compute the size of the maximum independent set, count the number of independent sets of a given size, and enumerate the independent sets of a given size.
%By making use of generic programming, our algorithms are very simple to implement and one can in addition directly utilize recent advances in tensor network contraction techniques such as near-optimal contraction order finding and slicing to achieve high performance.
By using generic programming techniques, %which require a program to work correctly for different input data types without sacrificing efficiency,
the same general, simple-to-implement framework can be used to compute all of these properties. Our algorithm utilizes recent advances in tensor network contraction techniques to achieve high performance, including methods to quickly find a near-optimal contraction order and slicing.
The algorithmic complexity of this approach is $2^{{\rm tw}(G)}$, where ${\rm tw} (G)$ is the treewidth of the problem graph $G$.
Other hard graph problems solvable in our framework include the maximum cut problem, the vertex coloring problem, the maximal clique problem, the dominating set problem, and the satisfiability problem, among others.
To demonstrate the versatility of this tool, we apply it to a few examples, including computing the entropy constant for several hardcore gases on disordered lattices, studying the overlap gap property on King's graphs at $0.8$ filling and 3-regular graphs, and visualizing the output of quantum optimization algorithms for the independent set problem. % including the calculations of the entropy constant for some hardcore lattice gases on 2D square lattices and the study of the overlap gap property on three regular graphs.
We show how our method helps to understand these hard problems better on systems with hundreds of vertices or spins. \xpurple{Say on systems with hundreds of vertices or spins?}
\end{abstract}

@@

The most studied aspect of the above problems is finding an optimal solution, 
%The most studied version of the above problems is finding an optimal solution maximum/minimum set satisfying the contraints and the corresponding set size,
, which is typically 
%while many of them belong to the hardness category
NP-hard~\cite{Hastad1996}, i.e. it is unlikely to be solved in a polynomial time.
%\purple{Is MIS really the most studied NP-hard optimization problem? I would guess it is heavily studied, but not necessarily the most studied.}
%\blue{Here, we haven't mentioned the independent set, here ``version'' means, e.g. counting, enumeration, devision. I think finding one of the largest/smallest set and its cardinality in the above mentioned problems is the most studied?}
In this paper, we focus on the less studied problem of computing \textit{solution space properties}  which can be much harder, e.g. counting independent sets of a general graph is \#P-complete~\cite{Vadhan2001}, but is crucial towards understanding a hard problem.
Here, the \textit{solution space property} refers to a class of quantities that not only include the maximum/minimum set size,
but also include the number of sets at a give size, enumeration of all sets at a give size or direct sampling of such sets when they are not enumeratable.
For the weighted version of the above problems, it also includes finding largest/smallest $k$ sets and their sizes.
%are many interesting and hard computational problems concerning various properties of independent sets.
Solution space property computation allows us to understand the solution space better.
With the counting of sets at different sizes, one can related the computational hardness with the counting of sets at different sizes~\cite{Xu2018}.
With the configuration enumeration/sampling at a given size, one can measure the overlap gap property~\cite{Gamarnik2013, Gamarnik2019} better,
while the configuration space connectivity also privides us an intuition about designing a better classical or quantum algortihms.
\purple{I would give examples of why we want to compute solution space
properties.
For instance, in statistical physics, computing thermal states and
observables via the partition function. In complexity theory, proving limits on local
algorithms (OGP).}\blue{now?}

@@

To avoid the discussion being unnecessarily diverged, we mainly focus on the independent set problem,
and discuss other combinatorial optimization problems,
including the cutting problem, the matching problem, the vertex coloring problem, the satisfiability problem, the dominating set problem, the set packing problem and the maximal clique problem, in \App{sec:otherproblems}.
%\xpurple{More details on which combinatorial optimization problems this applies to?}
The paper is organized as follows.
We first introduce the basic concepts of tensor networks and generic programming in \Sec{sec:tn} and \Sec{sec:generic}.
Then we show how to reduce the independent set problem to a tensor network contraction problem in \Sec{sec:tnmap}
and show how to engineer the element types in \Sec{sec:counting}, \Sec{sec:enumeration} and \Sec{sec:weighted} to count independent sets at a given size, enumerate/sample independent sets at a given size and find largest set sizes in weighted graphs.
% by computing a number of independent set properties of certain sparse graphs on central processing units (CPUs) and graphics processing units (GPUs) and show its good performance.
Lastly, we benchmark our algorithms in \Sec{sec:benchmark} and provide three examples in \Sec{sec:examples} to demonstrate the versatility of our tool.
In the first example, we show how our method can help computing the entropy constant for some hardcore lattice gases on 2D square lattices. We compute the 2D generalization of Fibonacci integer sequence up to size $39$, while the previous best known result is $37$.
In the second example, we analyze the pairwise Hamming distances from configurations directly sampled from large independent sets of King's graphs at $0.8$ filling and three regular graphs. We show the strongest numeric evidence so far of the absence of overlap gap property in King's graph and the existence of overlap gap properties in three regular graphs.
In the last example, we show how our algorithm helps people to understand the performance of a quantum variational algorithms on Rydberg atom arrays to find maximum independent sets. This is a research frontier that understand which can help people design better quantum algorithms to beat its classical counterpart.
%Our method can also be used to find ``maximal'' independent set properties; 
%a maximal independent set is an independent set that is not a subset of any other independent set, but its size may not be the maximum.

@@

\begin{definition}[Generic programming ~\cite{Stepanov2014}]
   Generic programming is an approach to programming that focuses on designing algorithms and data structures so that they work in the most general setting without loss of efficiency.
\end{definition}
This definition of generic programming covers two major aspects: ``work in the most general setting'' and ``with out loss of efficiency''.
By in the most general setting, we mean that a single program should work correctly for different input data types. For example, 
%\purple{I think the following paragraph could be clarified. When you read the definition of generic programming, it's not clear what ``in the most general setting'' means. My understanding is it just means that the program will work regardless of the input data type - it would be good to say this explicitly. }
%\blue{This definition is from the book I cited, I also do not like ``in the most general setting'', maybe we just use our version?}
%\purple{I also think the way it is written takes a while to understand what dynamically versus statically typed languages are. The definition, “When these languages “see” a new input type, the compiler can recompile the generic program for the new type.” comes rather late in the paragraph.}
%\xpurple{Also, while I know what you mean, concepts like “inlining immutable elements” and “cache miss rate” would be
%lost on someone only with a physics background.}
%This definition of generic programming contains two major aspects: a single program works in the most general setting and efficiency.
%To understand the first aspect on generality,
suppose we want to write a function that raises an element to a power, $f(x, n) := x^n$.
One can easily write a function for standard number types that computes the power of $x$ in $O \left( \log(n) \right)$ steps using the multiply and square trick.
Generic programming does not require $x$ to be a standard number type,
instead it treats $x$ as an element with an associative multiplication operation $\odot$ and a multiplicative identity $\mymathbb{1}$.
In such a way, when the program takes a matrix as an input, it computes the matrix power without extra efforts.
The second aspect is about the efficiency. For dynamically typed languages, such as Python,
one can easily write very general codes, but the efficiency is not guaranteed; for example, the speed of computing the matrix multiplication between two numpy arrays with python objects as elements is much slower than statically typed languages such as C++ and Julia~\cite{Bezanson2012}.
C++ uses templates for generic programming while Julia takes advantage of just-in-time compilation and multiple dispatch.
When these languages ``see'' a new input type, the compiler can recompile the generic program for the new type.
A myriad of optimizations can be done during the compilation, such as optimize the memory layout of immutable elements with fixed sizes in an array to speed up the array indexing.
In Julia, arrays with immutable elements with fixed sizes can even be compiled to GPU devices for faster computation~\cite{Besard2018}.

This motivates us to think about what is the most general tensor element type allowed in a tensor network contraction program.
We find that, as long as the algebra of tensor elements forms a commutative semiring, the tensor network contraction result will be valid and be independent of the contraction order.
A commutative semiring is a ring with its multiplication operation being commutative and without an additive inverse.
To define a commutative semiring with the addition operation $\oplus$ and the multiplication operation $\odot$ on a set $R$, the following relations must hold for any arbitrary three elements $a, b, c \in R$.
\begin{align*}
(a \oplus b) \oplus c = a \oplus (b \oplus c) & \hspace{5em}\text{$\triangleright$ commutative monoid $\oplus$ with identity $\mymathbb{0}$}\\
a \oplus \mymathbb{0} = \mymathbb{0} \oplus a = a &\\
a \oplus b = b \oplus a &\\
&\\
(a \odot b) \odot c = a \odot (b \odot c)  &   \hspace{5em}\text{$\triangleright$ commutative monoid $\odot$ with identity $\mymathbb{1}$}\\
a \odot  \mymathbb{1} =  \mymathbb{1} \odot a = a &\\
a \odot b = b \odot a &\\
&\\
a \odot (b\oplus c) = a\odot b \oplus a\odot c  &  \hspace{5em}\text{$\triangleright$ left and right distributive}\\
(a\oplus b) \odot c = a\odot c \oplus b\odot c &\\
&\\
a \odot \mymathbb{0} = \mymathbb{0} \odot a = \mymathbb{0}
\end{align*}
We require the algebra being multiplicative commutative because the contraction order optimizer will change the multiplication order of elements, while we require that the contraction order does not change the contraction results.
In the following sections, we show how to compute a number of properties of independent sets by designing tensor element types as specific commutative semirings without changing the tensor network contraction program~\cite{Stepanov2014}.
The Venn diagram in \Fig{fig:venn-diagram} shows the algebras we will introduce in the main text and how are they related, and Table~\ref{tbl:generictypes} summarizes those properties that can be computed by various tensor element types.

\begin{figure}[th]
   \centering
\centerline{\begin{tikzpicture}[]
    \scriptsize
    \node[draw,fill=lime!80,fill opacity=1, text opacity=1.0,ellipse,minimum width=2cm, minimum height=1cm,inner sep=0pt] at (0, 2.5) (R) {$\mathbbm{R}$};
    \def\dx{-3};
    \node[draw,fill=teal!50,fill opacity=1, text opacity=1.0,ellipse,minimum width=5cm, minimum height=3cm,inner sep=0pt] at (\dx, 0) (PN) {\hspace{-3.5cm}Polynomial};
    \node[draw,fill=brown!75,fill opacity=1, text opacity=1.0,ellipse,minimum width=3.5cm, minimum height=1.0cm,inner sep=0pt] at (\dx, 0.7) (P1) {\hspace{-1.0cm}Largest order};
    \node[draw,fill=brown!40,fill opacity=1, text opacity=1.0,ellipse,minimum width=3.5cm, minimum height=1.0cm,inner sep=0pt] at (\dx, -0.4) (P2) {\hspace{-0.8cm}Largest 2 orders};
    \node[draw,fill=brown,fill opacity=1, text opacity=1.0,ellipse,minimum width=0.8cm, minimum height=0.3cm,inner sep=0pt] at (\dx, 1.0) (T) {};
    \node[draw,fill=brown!50,fill opacity=1, text opacity=1.0,ellipse,minimum width=0.8cm, minimum height=0.3cm,inner sep=0pt] at (\dx, -0.1) (T2) {};
    \node at (\dx, -1.2) {$\ldots$};
    \node[above = 1cm] at (T) (textT) {Tropical};
    \node[below = 2cm, left=4cm] (textT2) {Extended Tropical};
    \draw[black,-latex] (textT) -- (T);
    \draw[black,-latex] (textT2) -- (T2);

    \caption{The the tensor network element types used in this work, while the purpose of these element types can be found in \Tbl{tbl:generictypes}.
    The overlap between two algebra indicates that a new algebra can be created by combining those two types of algebra. ``Largest order'' and ``Largest 2 orders'' mean truncating the polynomial by only keeping its largest or largest two order.}
     \label{fig:venn-diagram}
\end{figure}

\begin{table}[t!]\centering
\begin{minipage}{\columnwidth}
\ra{1.3}
        \begin{tabularx}{\textwidth}{sb}\toprule
            \hline
   \textbf{Element type}     & \textbf{Solution space property} \\
   {$\mathbbm{R}$}     & {Counting of all independent sets} \\
   {Polynomial} (\Eq{eq:polynomial})     & {Independence polynomial} \\
   {Tropical (\Eq{eq:tropical})}    & {Maximum independent set size} \\
   {Extended tropical of order $k$ (\Eq{eq:ext-tropical})}    & {Largest $k$ independent set sizes} \\
   {Polynomial truncated to $k$-th order (\Eq{eq:countingtropical} and \Eq{eq:max2poly})}     & {$k$ largest independent sizes and their degeneracy} \\
   {Set} (\Eq{eq:set})     & {Enumeration of independent sets} \\
   {Expression tree} (\Eq{eq:expr})     & {Sampling of independent sets} \\
   {Polynomial truncated to largest order combined with bit string} (\Eq{eq:singleconfig})     & {Maximum independent set size and one of such configurations} \\
   {Polynomial truncated to $k$-th order combined with set} (\Eq{eq:countingtropicalset})    & {$k$ largest independent set sizes and their enumeration} \\
            \bottomrule
        \end{tabularx}
    \caption{Tensor element types and the independent set properties that can be computed using them.}\label{tbl:generictypes}
\end{minipage}
\end{table}

@@

The contraction of this tensor network can be done in a pairwise order utilizing the associativity, additive commutativity, and multiplicative commutativity of tensor elements. The contraction process can be represented as a binary tree structure:

\centerline{\begin{tikzpicture}[]
    \def\dx{0};
    \def\dy{0};
    \def\d{1};
    \def\a{1.0};
    \def\b{1};
    \def\ya{-1.0};
    \def\yb{-0.8};
    \node[] at (\dx+5*\a, \dy+5*\b) (R1) {$R[n^2]$};
    \node[] at (\dx+3*\a, \dy+4*\b) (S1) {$S_{bc}[n^3]$};
    \node[] at (\dx+2*\a, \dy+3*\b) (Q1) {$Q_{bd}[n^2]$};
    \node[] at (\dx+8.5*\a, \dy+3*\b) (Q2) {$Q_{bc}[n^2]$};
    \node[] at (\dx+\a, \dy+2*\b) (P1) {$P_{d}[n]$};
    \node[] at (\dx+9*\a, \dy+2*\b) (P2) {$P_{bc}[n^3]$};
    \node[] at (\dx+0.5*\d, \dy+\b) (M1) {$M_{d}[n^2]$};
    \node[] at (\dx+3.5*\d, \dy+\b) (M2) {$M_{bd}[n^2]$};
    \node[] at (\dx+5.5*\d, \dy+\b) (M3) {$M_{cd}[n^2]$};
    \node[] at (\dx+9.5*\d, \dy+\b) (M4) {$M_{ac}[n^2]$};
    \node[] at (\dx, \dy) (Bde) {$B_{de}$};
    \node[] at (\dx+\d, \dy) (We) {$W_e$};
    \node[] at (\dx+2*\d, \dy) (Wd) {$W_d$};
    \node[] at (\dx+3*\d, \dy) (Bbd) {$B_{bd}$};
    \node[] at (\dx+4*\d, \dy) (Wb) {$W_b$};
    \node[] at (\dx+5*\d, \dy) (Bcd) {$B_{cd}$};
    \node[] at (\dx+6*\d, \dy) (Wc) {$W_c$};
    \node[] at (\dx+7*\d, \dy) (Bbc) {$B_{bc}$};
    \node[] at (\dx+8*\d, \dy) (Bab) {$B_{ab}$};
    \node[] at (\dx+9*\d, \dy) (Bac) {$B_{ac}$};
    \node[] at (\dx+10*\d, \dy) (Wa) {$W_a$};
    \draw[] (Bde) -- (M1);
    \draw[] (We) -- (M1);
    \draw[] (Bbd) -- (M2);
    \draw[] (Wb) -- (M2);
    \draw[] (Bcd) -- (M3);
    \draw[] (Wc) -- (M3);
    \draw[] (Bac) -- (M4);
    \draw[] (Wa) -- (M4);
    \draw[] (M1) -- (P1);
    \draw[] (Wd) -- (P1);
    \draw[] (Bab) -- (P2);
    \draw[] (M4) -- (P2);
    \draw[] (P1) -- (Q1);
    \draw[] (M2) -- (Q1);
    \draw[] (Bbc) -- (Q2);
    \draw[] (P2) -- (Q2);
    \draw[] (Bbc) -- (Q2);
    \draw[] (P2) -- (Q2);
    \draw[] (Q1) -- (S1);
    \draw[] (M3) -- (S1);
    \draw[] (S1) -- (R1);
    \draw[] (Q2) -- (R1);
\end{tikzpicture}}
% \small{
% \begin{align*}
%     &\sum_{s_a,s_b,s_c,s_d,s_e} W(x_a, w_a)_{s_a} W(x_b, w_b)_{s_b} W(x_c, w_c)_{s_c} W(x_d, w_d)_{s_d} W(x_e, w_e)_{s_e} B_{s_a s_b} B_{s_b s_d} B_{s_c s_d} B_{s_a s_c} B_{s_b s_c} B_{s_d s_e}\\
%     =&\sum_{s_b,s_c}\left(\sum_{s_d}\left(\left(\left(\left(\sum_{s_e}B_{s_d s_e}W(x_e,w_e)_{s_e}\right) W(x_d,w_d)_{s_d}\right) \left(B_{s_bs_d} W(x_b, w_b)_{s_b}\right)\right) \left(B_{s_cs_d} W(x_c, w_c)_{s_c}\right)\right)\right.\\
%     &\phantom{XXX}\left.\left(B_{s_bs_c}\left(\sum_{s_a}B_{s_as_b}\left(B_{s_as_c}W(x_a,w_a)_{s_a}\right)\right)\right)\right)\\
%     =&1 + x_a^{w_a} + x_b^{w_b} + x_c^{w_c} + x_d^{w_d} + x_e^{w_e} + x_a^{w_a}x_d^{w_d} + x_a^{w_a}x_e^{w_e} + x_c^{w_c}x_e^{w_e} + x_b^{w_b}x_e^{w_e}
%     %=&1+5x+4x^2 \qquad \quad (x_{i} = x, w_{i}=1).
% \end{align*}}
The contraction process goes from bottom to top, where the root stores the contraction result that we want, the leaves are tensors in the our tensor network while the rest are all intermediate contraction results. Tensor subscripts are indices so that the number of subscripts indicates the space complexity to store this tensor, and the contraction complexity to generate a tensor is annotated in the square brackets, where $n$ is the dimension of degrees of freedom, which in our case is always $2$.
One can easily check the largest tensor during computing has space complexity $n^2$, and the largest contraction time complexity is $n^3$, which is much smaller than direct enumeration ($n^5$).
\end{example}

@@

and then apply the Gaussian elimination to fit the coefficients of this polynomial.

@@

This algebra is a commutative semiring because we define the equivalence of two expression trees by comparing their expanded (using \Eq{eq:set}) forms, rather than their storages.
Although one is probably unable to enumerate the configurations represented by this tree due to the extremely large configuration space size,
but he can directly sample configurations from this tree.
This is because once we have this expression tree representation of the configuration space,
either generate by using this algebra directly or serve as the coefficients of (truncated) polynomials,
one can compute the number of configurations in each branch rigorously and efficiently.
%\purple{To clarify - is the expression tree now the coefficients
%on the polynomial? I am a little confused on how we would, for instance, use this to
%compute the MISs}\blue{added one more descriptive sentence, does it resolve your concern?}
To generate samples, one just descends this expression tree to left or right sibling with probability decided by the size of each branch while doing the computation specified by the first field.
\xpurple{How do we compute the size of each
branch?}
\begin{example}
Let us consider the following expression tree 
\begin{equation*}
    (\texttt{SUM}, /, (\texttt{PROD}, /, A, B), (\texttt{SUM}, /, C, D)),
\end{equation*}
where siblings $A, B, C$ and $D$ can be any four types of nodes.
It can be represented diagrammatically as the following.

@@

We compare two categories of graphs, one is the King's graph at $0.8$ filling,

@@

Which is a clear evidence of that the overlap gap property is absent in $0.8$ filling King's graph.
Given the fact that the MIS problem on an arbitrary graph can be mapped to a King's graph at a certain filling~\cite{Ebadi2022}, this result is highly nontrivial.

@@
